Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-30 2024/OPTIMA/OPTIMA_2024_MCO.tex
last section

In the first experiment 100 bi-criterial MCO problems have been solved. As the partial criteria, the functions \cite{ML_MCO_2023} 
\begin{equation}
    \label{eq:19}
		\begin{matrix}
		  f(y)= -(AB + CD)^{1/2} \\
			AB =(\sum_{i=1}^7{\sum_{j=1}^7{[A_{ij} a_{ij} (y_1,y_2) + B_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			CD =(\sum_{i=1}^7{\sum_{j=1}^7{[C_{ij} a_{ij} (y_1,y_2) - D_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			a_{ij} (y_1,y_2) = \sin(\pi i y_1) \sin(\pi j y_2), \\
			b_{ij} (y_1,y_2) = \cos(\pi i y_1) \cos(\pi j y_2),
		\end{matrix}
\end{equation}
were taken  where $0 \leq y_1, y_2 \leq 1$, parameters $-1 \leq A_{ij},B_{ij},C_{ij},D_{ij} \leq 1$ are independent uniformly distributed random values.

Computational experiments for techniques of Subsection \ref{ssec:42} which utilize probabilities of belonging to the Pareto set can be divided into 3 groups.
\begin{enumerate}
	\item 	Experiments with the algorithm ML\_MGSA\_Prob (fixed weights).
	\item 	Solving MCO problems by the method ML\_MGSA\_Prob (Adjustive weights));
	\item 	Experiments with changing weights in the course of ML\_MGSA\_Prob running and transformed probabilities (\ref{eq:18})  (ML\_MGSA\_Prob (Norm Log)).
\end{enumerate}

In each series of experiments for these optimization algorithms, the coefficients $\alpha$ from (\ref{eq:16}) changed in the range $[0.1, 0.9]$. Comparison with the algorithm MGSA that does not use machine learning methods (in the characteristic (\ref{eq:16}) $\alpha = 0$) and with the algorithm ML\_MGSA\_Dist taking into account the distance to the separating hyperplane was carried out. The comparison was made on the base of  the calculated HV index characterizing quality of the obtained Pareto set approximation \cite{ML_MCO_2023,Evtushenko2014}. Further subsections include the results of only those calculations in which the $\alpha$ parameter provided the best quality of the solution found. Tables contain results of solving all the problems (\ref{eq:07}) with fixed accuracy $\varepsilon=0.01$ in the termination condition (\ref{eq:14}). Figures present results for a series of decreasing $\varepsilon$ which generates increasing numbers of trials.

@@

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[ht]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup iter.} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

@@

Acceleration of the algorithm (in terms of reducing the number of search trials required to evaluate the Pareto set) is achieved through the application of machine learning methods. Namely, when using the SVM method, trial points in the criteria space are classified (each point is/is not an effective solution), on the basis of which a numerical estimate of the probability of belonging to the Pareto set is built. Polynomial and radial basis function kernels in SVM allows one to adequately approximate Pareto front of a complex structure.

Solving several series of MCO problems with complex multiextremal criteria has demonstrated that the proposed approach to evaluating the set of effective solutions is promising. Further research aims to develop the proposed approach by consideration of other machine learning models (in particular, artificial neural networks) to better classify trials and approximate the Pareto set.

>>>>

In the first experiment 100 bi-criterial MCO problems have been solved. As the partial criteria, the functions $f(y)$ from \cite{ML_MCO_2023} were taken,
\begin{equation}
    \label{eq:19}
		\begin{matrix}
		  f(y)= -(AB + CD)^{1/2} \\
			AB =(\sum_{i=1}^7{\sum_{j=1}^7{[A_{ij} a_{ij} (y_1,y_2) + B_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			CD =(\sum_{i=1}^7{\sum_{j=1}^7{[C_{ij} a_{ij} (y_1,y_2) - D_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			a_{ij} (y_1,y_2) = \sin(\pi i y_1) \sin(\pi j y_2), \\
			b_{ij} (y_1,y_2) = \cos(\pi i y_1) \cos(\pi j y_2),
		\end{matrix}
\end{equation}
where $0 \leq y_1, y_2 \leq 1$ and parameters $-1 \leq A_{ij},B_{ij},C_{ij},D_{ij} \leq 1$ are independent uniformly distributed random values.

Computational experiments for proposed techniques which utilize probabilities of belonging to the Pareto set can be divided into 3 groups.
\begin{enumerate}
	\item 	Experiments with the algorithm ML\_MGSA\_Prob (fixed weights);
	\item 	Solving MCO problems by the method ML\_MGSA\_Prob (Adjustive weights);
	\item 	Experiments with changing weights in the course of ML\_MGSA\_Prob running and transformed probabilities (\ref{eq:18})  (ML\_MGSA\_Prob (Norm Log)).
\end{enumerate}

In each series of experiments for these optimization algorithms, the coefficients $\alpha$ from (\ref{eq:16}) changed in the range $[0.1, 0.9]$. Comparison with the algorithm MGSA that does not use machine learning methods (i.e. with $\alpha = 0$ in the characteristic (\ref{eq:16})) and with the algorithm ML\_MGSA\_Dist taking into account the distance to the separating hyperplane was carried out. The comparison was made on the base of  the HV index characterizing quality of the obtained Pareto set approximation \cite{ML_MCO_2023,Evtushenko2014}. Further subsections include the results of only those calculations in which the $\alpha$ parameter provided the best quality of the solution found. Tables contain results of solving all the problems (\ref{eq:07}) with fixed accuracy $\varepsilon=0.01$ in the termination condition (\ref{eq:14}). Figures present results for a series of decreasing $\varepsilon$ which generates increasing numbers of trials.

@@

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[ht]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

@@

Acceleration of the algorithm (in terms of reducing the number of search trials required to evaluate the Pareto set) is achieved through the application of machine learning methods. Namely, when using the SVM method, trial points in the criteria space are classified (each point is/is not an efficient solution), on the basis of which a numerical estimate of the probability of belonging to the Pareto set is built. Polynomial and radial basis function kernels in SVM allows one to adequately approximate Pareto front of a complex structure.

Solving several series of MCO problems with complex multiextremal criteria has demonstrated that the proposed approach to evaluating the set of efficient solutions is promising. Further research aims to develop the proposed approach by consideration of other machine learning models (in particular, artificial neural networks) to better classify trials and approximate the Pareto set.
